Could tides explain the extended stellar densities of the Sculptor and Ursa Minor dwarfs?

### 2025-11-13 paper/paper.tex
paper updates

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% if you use custom commands in your title,
% ensure to check your title when submitting!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{The extended stellar densities of the Sculptor and Ursa Minor dwarf galaxies}

\subtitle{Innate nature or tidal nurture?}

@@

        \institute{%
            Department of Physics \& Astronomy, University of Victoria, Finnerty Road, Victoria, British Columbia V8A 1A1, Canada
             \email{danielaboyea@gmail.com}
             \thanks{Shows the usage of elements in the institute field}
            \and others\\ }

@@

  \abstract
  % context heading (optional)
  % {} leave it empty if necessary  
   {Most dwarf spheroidals (dSph) of the Milky Way follow an exponential surface density profile. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We aim to determine if tides may explain the excess of stars in Sculptor and Ursa Minor.} 
  % methods heading (mandatory)
   {We develop N-body models simulating tidal effects in a Milky Way and Large Magellanic Cloud (LMC) potential. }
  % results heading (mandatory)
   {Neighet galaxy is strongly tidally affected. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component barely evolves. UMi loses about 90\% of its dark matter mass but the stellar component would not be observationally affected. }
  % conclusions heading (optional), leave it empty if necessary
   {We conclude that the excess of stars likely reflects a divergent assembly history as compared to other local dSphs, possibly explained by past mergers or star formation bursts. }

   \keywords{giant planet formation --
                $\kappa$-mechanism --
                stability of gas spheres
               }

@@

Since \citet{faber+lin1983}, exponential density profiles have been commonly used for dwarf
spheroidal galaxies \citealt{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
However, Sculptor and Ursa Minor, among Milky Way satellites, show strong deviations from an exponential profile (figure X), often speculated to be a result of tidal interactions
\citep[e.g.,][]{innanen+papp1979, exkridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003} or sometimes a ``halo''of stars
surrounding the dwarf \citep{westfall+2006}.
Most recently, \citet{sestito+2023a, sestito+2023b} report a
``kink'' in the density profile of each galaxy. They spectroscopically
follow up distant stars, finding members as far as 6 and 12 half-light
radii from the centre of each dwarf. If these dwarfs had exponential
profiles, like Fornax, then these far-outlying stars should be much
rarer.

The Local Group hosts several examples of ongoing tidal disruption. The
Magellanic stream, a massive, gas-rich feature emanating from the
Magellanic clouds, is believed to arise partially from the MW's tides
\citep{putman+1998, diaz+bekki2012, donghia+fox2016}. Other clear
examples of tidal streams include the Sagittarius stream, the Andromeda
Giant Southern stream, and the Tucana III stream
\citep[e.g.,][]{ibata+gilmore+irwin1994, ibata+2001, li+2018}. These
examples illustrate that hierarchical accretion remains an active
process. Interpreting such observations relies on simulations of tidal
disruption.

Cosmological simulations struggle to resolve tidal effects on dwarfs.
Since many dwarfs are near the resolution limit, they are vulnerable to
artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
numerical challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.

Idealized simulations predict clear properties of tidally disrupting
dwarf spheroidal galaxies. Tidally stripped stars form \emph{tidal
streams}---stellar tails with a bulk velocity gradient
\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
Most mass loss happens near pericentre, where tides are strongest.
However, the central structure of a dwarf galaxy often remains
undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. For instance,
NFW halos are also found to be resilient to full tidal disruption
\citep{EP2020}, but cored dark matter halos may disrupt fully and faster
\citep[e.g.,][]{penarrubia+2010, errani+2023a}.
To first order, tidal mass loss peels away the outer layers of a dwarf
galaxy in energy space.
\citet{drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}
showed that tidal effects are nearly entirely described as the removal
of particles above a truncation energy \citep[see
also][]{choi+weinberg+katz2009}. \citet{stucker+2023} generalized this
idea, creating a model for adiabatic tidal mass loss in an isotropic
tidal field. Their model explains the resilience of NFW halos against
full tidal disruption and the origin of well-defined ``tidal tracks''
\citep[as observed in][]{PNM2008, green+vandenbosch2019, EN2021}.

With precise orbital constraints and improved models of the Milky Way
potential, recent studies have continued to probe the dynamical
histories of individual dwarf galaxies.
\citet{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024}
both ran simulations tuned to Fornax, showing that this galaxy's stellar
component or globular clusters are likely not affected by tides.
Similarly, \citet{borukhovetskaya+2022a} analyzed Crater II, showing
that the present-day structure is challenging to reconcile with NFW
initial conditions and Galactic tides. Most relevantly,
\citet{iorio+2019} also tailored simulations to Scl, finding weak
Galactic tidal influence.

The \textbf{Jacobi radius} represents the approximate radius where stars
become unbound for a galaxy in a circular orbit around a host
galaxy.\footnote{The Jacobi radius was derived at least as early as
  \citet{laplace1798}. This radius also bears other names, such as the
  Hill radius \citep[from][]{hill1878}. Likely only named after Jacobi
  for the Jacobi integral \citep{jacobi1836}.} Calculated from an
approximation of the location of the \(L_1\) and \(L_2\) Lagrange
points, the Jacobi radius is where the mean density of the dwarf galaxy
is roughly three times the mean interior density of the host galaxy at
pericentre, or
\begin{equation}\protect\phantomsection\label{eq:r_jacobi}{
3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
}\end{equation} \citep[ eq. 7-84]{BT1987}. If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. While strictly valid for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages.

We also use the \textbf{break radius} as defined in
\citet{penarrubia+2009}, marking the outermost radius within which the
dwarf has been able to achieve equilibrium after pericentric passage in
a highly eccentric orbit. The break radius \(r_{\rm break}\) is
proportional to the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\),
\begin{equation}\protect\phantomsection\label{eq:r_break}{
r_{\rm break} = C\,\sigma_{v}\,\Delta t
}\end{equation} where the empirical constant is \(C \approx 0.55\).
\(r_{\rm break}\) describes where the dynamical timescale becomes longer
than the time since the perturbation, i.e., the radius within which the
galaxy is dynamically relaxed.

Building on this body of work, we will use idealized simulations to
understand the severity of tidal effects on Sculptor and Ursa Minor.

\begin{table}[t]
\centering
\caption[Observed Properties of Sculptor]{Observed properties of Sculptor. References are: (1) Muñoz et al. (2018, Sérsic fit), (2) Tran et al. (2022, RR lyrae distance), (3) Alan W. McConnachie and Venn (2020b), (4) Arroyo-Polonio et al. (2024). }
\label{tbl:scl_obs_props}
\begin{tabular}{lll}
\hline
parameter & value & Source\\
\hline
$\alpha / ^\circ$ & $15.0183 \pm 0.0012$ & 1\\
$\delta / ^\circ$ & $-33.7186 \pm 0.0007$ & ”\\
distance modulus & $19.60 \pm 0.05$ & 2\\
distance & $83.2 \pm 2$ kpc & ”\\
$\mu_{\alpha*} / \masyr$  & $0.099 \pm 0.002 \pm 0.017$ & 3\\
$\mu_\delta / \masyr$ & $-0.160 \pm 0.002_{\rm stat} \pm 0.017_{\rm sys}$  & ”\\
$\V_{\rm los}$ / ${\rm km\,s^{-1}}$ & $111.2 \pm 0.3\ $ & 4\\
$\sigma_v$ / ${\rm km\,s^{-1}}$ & $9.7\pm0.2\ $ & ”\\
$R_h/'$ & $9.79 \pm 0.04$ & 1\\
ellipticity & $0.37 \pm 0.01$ & ”\\
position angle & $94\pm1^\circ$ & ”\\
$M_V$ & $-10.82\pm0.14$ & ”\\
\hline
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption[Observed Properties of Ursa Minor]{Observed properties of Ursa Minor. References are: (1) Muñoz et al. (2018, Sérsic fit), (2) Garofalo et al. (2025, RR lyrae distance), (3) Alan W. McConnachie and Venn (2020a), (4) Pace et al. (2020), average of MMT and Keck results with systematic uncertainty based on the difference between the MMT and Keck means. }
\label{tbl:umi_obs_props}
\begin{tabular}{lll}
\hline
parameter & value & Source\\
\hline
$\alpha/^\circ$ & $ 227.2420 \pm 0.0045$ & 1\\
$\delta/^\circ$ & $67.2221 \pm 0.0016$ & 1\\
distance modulus & $19.23 \pm 0.11$ & 2\\
distance/kpc & $70.1 \pm 3.6$ & ”\\
$\mu_\alpha*$ / mas yr$^{-1}$  & $-0.124 \pm 0.004 \pm 0.017$ & 3\\
$\mu_\delta$ / mas yr$^{-1}$  & $0.078 \pm 0.004_{\rm stat} \pm 0.017_{\rm sys}$ & ”\\
$\V_{\rm los}$ / km s$^{-1}$  &$-245.9 \pm 0.3_{\rm stat} \pm 1_{\rm sys}$ & 4\\
$\sigma_v$ & $8.6 \pm 0.3$ & 4\\
$R_h$ & $11.62 \pm 0.1$ & 1\\
ellipticity & $0.55 \pm 0.01$ & 1\\
position angle & $50 \pm 1^\circ$ & 1\\
$M_V$ & $-9.03 \pm 0.05$ & 1\\
\hline
\end{tabular}
\end{table}

@@

Reducing background-contamination is essential to determine faint structures around dwarf galaxies. Before {\it Gaia}, most studies relied on colour-magnitude selections to select stars more likely to be members. Now with {\it Gaia}, proper motions and parallax information can substantially reduce the background of likely stars, producing low-contamination membership lists for dwarf galaxies.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data. J+24 used a Bayesian framework
incorporating proper motion (PM), colour-magnitude diagram (CMD), and
spatial information to determine the probability that a given star
belongs to the satellite or foreground/background. By accounting for PM
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component. J+24 detected
candidate members out to \textasciitilde10 half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

For our fiducial sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\). We do not filter on magnitudes explicitly, but
J+24's quality cuts typically only include stars with \(G < 21\). We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. Most stars have \(P_{\rm  sat}\) values which are nearly 0 or
1, so the exact choice of probability threshold has little effect on the
resulting sample. Even at our relatively generous probability threshold
of 0.2, the purity remains high when validated against spectroscopic
line-of-sight (LOS) velocities (\textasciitilde90\%, J+24).\footnote{This
  would indicate that the J+24 model probabilities are mis-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate.} However, we
find that our conclusions are unchanged when limiting samples to only
the brightest stars. Altogether, the J+24 method provides a
high-quality, low-contamination sample of dwarf galaxy candidate member
stars, which we will now investigate in further detail.

Fig.~\ref{fig:observed_density_profiles} compares the fiducial density
profiles of Sculptor, Ursa Minor, Fornax, and other classical dwarf
galaxies. Of the classicals, we exclude Antlia II, due to the extremely
high background, and Sagittarius, which was not included in J+24. The
density profiles are scaled to match at the half-light radius, taken
from \citet{munoz+2018}. All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{Comparing
  density profiles of dwarf galaxies is complicated by variations in the
  effective depth between galaxies. However, the deviations between Scl,
  UMi, and other galaxies are apparent even where the data is complete
  across all dwarfs.} In the outer regions, however, Sculptor and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards,
and at \(\sim 8 R_h\), may reach 2 orders of magnitude.

\begin{figure}
    \centering
    \includegraphics{figures/classical_dwarf_profiles.pdf}
    {\caption{The density profiles of classical dwarfs, using data from \citet{jensen+2024}.}}
    \label{fig:observed_density_profiles}
\end{figure}

\section{Methods}

\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
To convert observed positions and velocities to Galactocentric
coordinates, we use the Astropy v4 Galactocentric frame
\citep{astropycollaboration+2022}. Our Cartesian Galactocentric
coordinates here assume the Galactic centre is at
\([x, y, z] = [0,0,0]\), where \(x\) is the direction from the sun to
the Galactic centre, \(y\) is the direction of the motion of the Local
Standard of Rest, and \(z\) is the direction perpendicular to the
Galactic plane. The coordinate frame is also right-handed, such that the
\(z\)-angular momentum of the sun is negative (since the sun is at
\(x<0\)). In this frame, the solar position is
\([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
\citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
km/s
\citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
The uncertainties in this reference frame are typically smaller than the
uncertainties on a dwarf galaxy's distance and tangential velocity.

We adopt the Milky Way potential described in \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The Galactic bulge is described by a \citet{hernquist1990} potential,
\begin{equation}{
    \Phi(r) = - \frac{GM}{r + a},
}\end{equation} 
where \(a=1.3\,{\rm kpc}\) is the scale radius and
\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
disks are represented with the \citet{miyamoto+nagai1975} cylindrical
potential, 
\begin{equation}{
    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
}\end{equation} 
where \(a\) is the disc radial scale length, \(b\) is
the scale height, and \(M\) is the total mass of the disk. For the thin
disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).
The halo is an NFW \citep{NFW1996, 1997} dark matter halo (Eq.~\ref{eq:nfw}) 
\begin{equation}\protect\phantomsection\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and density. We parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\), or
\(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).

@@

To explore the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables. The present-day position,
distance modulus, LOS velocity, and proper motions are each sampled from
normal distributions given the reported uncertainties in
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
each sampled position/velocity back in time for 10 Gyr using \agama{}
\citep{agama}. Dynamical friction is not expected to impact orbits
substantially, so we assume a single point-mass particle for the backwards
integration.
Figs.~\ref{fig:scl_orbits},~\ref{fig:umi_orbits} shows 100 samples of these orbits.

To maximize tidal effects, we select an orbit with the \(\sim 6\sigma\)
smallest pericentre among all possible orbits integrated backwards for
\(10\,\Gyr\). We achieve this by taking the median parameters of all
orbits with a pericentre less than the 0.0027th quantile pericentre,
yielding a pericentre of 43 kpc. Given the current observations, it is
unlikely that Sculptor has had a significantly smaller pericentre than
our selected orbit, which we refer to as the \smallperi{} orbit. We take
the first apocentre after a look-back time of 10 Gyr, or at
\(\sim9.1\,\Gyr\), as the initial conditions for our model of Sculptor,
noted in Table~\ref{tbl:orbit_ics}.

We iteratively adjust Ursa Minor's initial
conditions to better match the present-day position. Initially, starting with low-resolution runs, we adjust the
cylindrical actions of the initial orbit by the final difference in
actions at the end of orbital evolution. After the initial actions have
converged (2 iterations), we change the initial action angles by the
final difference in action angles. This method converges within 4
iterations to an orbit agreeing with the observed kinematics of Ursa
Minor. Since Sculptor's orbit is less strongly affected by tides, we do
not carry out this correction for Sculptor.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% if you use custom commands in your title,
% ensure to check your title when submitting!
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Could tides explain the extended stellar densities of the Sculptor and Ursa Minor dwarfs?}

@@

        \institute{%
            Department of Physics \& Astronomy, University of Victoria, Finnerty Road, Victoria, British Columbia V8A 1A1, Canada
             \email{danielaboyea@gmail.com}
             %\thanks{Shows the usage of elements in the institute field}
            \and others\\ }

@@

  \abstract
  % context heading (optional)
   {Most dwarf spheroidals (dSph) of the Milky Way have light profiles following an exponential law, with a sharply declining outer density. However, Sculptor (Scl) and Ursa Minor (UMi) dSphs show an excess of stars at large radii.}
  % aims heading (mandatory)
   {We investigate whether tides may produce these extended light profiles.}
  % methods heading (mandatory)
   {We conduct idealized N-body simulations of both galaxies in the tidal field of the Milky Way and Large Magellanic Cloud (LMC). }
  % results heading (mandatory)
   {Tides affect neither galaxy strongly. Because the LMC perturbs Scl's orbit, Scl has few pericentres and the stellar component evolves little. UMi instead loses about 95\% of its dark matter mass, but the stars would still appear unaffected with current observations. }
  % conclusions heading (optional), leave it empty if necessary
   {The outer stellar excess in Scl and UMi are likely not of tidal origin. Instead, these features are likely innate, possibly explained by past mergers, accretion, or star formation episodes. }

   \keywords{Galaxies: dwarf --
       Galaxies: kinematics and dynamics --
       Local Group
               }

@@

Since \citet{faber+lin1983}, exponential surface density profiles have been commonly used to describe dwarf spheroidal (dSph) galaxies \citep[e.g.,][]{binggeli+sandage+tarenghi1984, mateo1998, mcconnachie+irwin2006, cicuendez+2018, kowalczyk+2013, martin+2016, MV2020a, battaglia+2022}.
However, the Sculptor (Scl) and Ursa Minor (UMi) dSphs show clear deviations from an exponential profile (Fig.~\ref{fig:observed_density_profiles}; e.g., \citealt{eskridge1988, IH1995, martinez-delgado+2001, walcher+2003, palma+2003}). This excess of stars at large radii was often speculated to be tidal interaction \citep[e.g.][]{innanen+papp1979, walcher+2003} or sometimes a `halo'of stars surrounding the dwarf \citep{westfall+2006}.
Most recently, \citet{sestito+2023a, sestito+2023b} report a
`kink' in the density profile of each galaxy. They spectroscopically
confirm distant members out to ${\sim}10$ half-light
radii from the centre of each dwarf. If these dwarfs had exponential
profiles, then these far-outlying stars should be much
rarer. We aim to test whether tides may form the extended stellar densities of Scl and UMi.

Cosmological simulations struggle to resolve tidal effects on dwarfs. Dwarfs are 
often near the resolution limit and vulnerable to artificial disruption
\citep[e.g.,][]{vandenbosch+2018, santos-santos+2025}. To overcome
these challenges, idealized simulations model a single subhalo in an
analytic host potential, achieving excellent numerical convergence \citep[e.g.,][]{hayashi+2003, bullock+johnston2005, klimentowski+2009, ogiya+2019}.
From these simulations, tidally stripped stars form \emph{tidal
streams}---stellar tails with a bulk velocity gradient
\citep[e.g.,][]{moore+davis1994, johnston+spergel+hernquist1995, read+2006}.
Most mass loss happens near pericentre, where tides are strongest.
However, the central structure of a dwarf galaxy often remains
undisturbed \citep{oh+lin+aarseth1995, piatek+pryor1995}. 
To first order, tidal mass loss peels away the outer layers of a dwarf
galaxy in energy space (\citealt{choi+weinberg+katz2009, drakos+taylor+benson2020, drakos+taylor+benson2022, amorisco2021}; or more generally, the adiabatic removal of stars on unbound orbits in \citealt{stucker+2023}). The most loosely bound particles are removed first---substantial dark matter mass can be removed before affecting the stellar component.

Improved orbital constraints and Milky Way
potential models have enabled precise dynamical modeling of dwarf galaxies.
Simulations tuned to Fornax \citep{battaglia+sollima+nipoti2015, borukhovetskaya+2022, dicintio+2024},
Scl \citep{iorio+2019}, or MW classicals\footnote{We define classicals to be dwarfs with absolute magnitudes between $-7.7$ and $-14$.} in general \citep{tchiorniy+genina2025} often find tides play a minor role in the stellar evolution of dwarf classicals. However, modeling tidal evolution under the Large Magellanic Cloud (LMC) has not yet been considered for most classical dSphs.

Two radii provide estimates of where tidal effects occur.
The \textbf{Jacobi radius}, $r_{\rm J}$. approximates the radius where stars
become unbound in a galaxy\footnote{ 
    While strictly valid only for circular orbits, assuming
\(r_{\rm peri}\) for the host-dwarf distance works as most stars are
lost during pericentric passages. $r_{\rm J}$ solves
}
\begin{equation}\label{eq:r_jacobi}
    3\bar \rho_{\rm MW}(r_{\rm peri}) \approx \bar \rho_{\rm dwarf}(r_J),
\end{equation}
given the mean enclosed  density of the host at pericentre $\bar \rho_{\rm MW}(r_{\rm peri})$,
and the mean enclosed satellite density $\bar \rho_{\rm dwarf}(r)$ \citep[eq. 7-84]{BT1987}.
If \(r_J\) is comparable to
the visible extent of a galaxy, we should expect to find clear signs of
tidal disturbance. 
The \textbf{break radius}, as defined in
\citet{penarrubia+2009}, marks the outermost dynamically-relaxed radius after pericentric passage. Based on the velocity dispersion, \(\sigma_v\), and time elapsed
since pericentre, \(\Delta t\), the break radius is
\begin{equation}\label{eq:r_break}
    r_{\rm break} = C\,\sigma_{v}\,\Delta t
\end{equation} 
where the empirical constant is \(C \approx 0.55\).

Building on this body of work, we will use idealized simulations to
understand the severity of tidal effects on Scl and UMi.

@@

Measuring the light profile of a resolved galaxy requires careful
consideration of whether any given star belongs to the system or not.
Without removing contamination from foreground/background sources, faint
features may be lost in the noise or be of uncertain association. In addition to colours and magnitudes, \emph{Gaia} now \citep{gaiacollaboration+2023} provides stellar parallax and proper motion which improve membership assignment.

Here, we use the \citepos[hereafter J+24]{jensen+2024} membership
probabilities from \emph{Gaia} data.\footnotemark{} J+24 used a Bayesian framework
incorporating proper motion, colour--magnitude diagram, and
spatial information to determine the probability that a given star
belongs to the satellite, $P_{\rm sat}$. By accounting for proper motions
in particular, J+24 produced low contamination samples of candidate
member stars out to large distances from a dwarf galaxy. J+24 extended
the algorithm presented in \citet{MV2020a, MV2020b} by additionally
including a secondary, extended spatial component, J+24 detected
candidate members out to ${\sim}10$ half-light radii from the
centres of some galaxies (\(R_h\)). Similar recent work includes
\citet{pace+li2019}; \citet{battaglia+2022}; \citet{pace+erkal+li2022};
\citet{qi+2022}.

\footnotetext{
We use
the \(P_{\rm sat}\) values from the elliptical 2-component runs if a
galaxy shows evidence for an outer component, the 1-component run
otherwise. 
}

For our sample, we adopt a minimum probability of
\(P_{\rm sat} = 0.2\).
%We do not filter on magnitudes explicitly, but J+24's quality cuts typically only include stars with \(G < 21\).
Most stars have \(P_{\rm  sat}\) values which are nearly 0 or 1, so the exact
choice of probability threshold has little effect on the resulting sample. Even
at the probability threshold of 0.2, the purity remains high when validated
against spectroscopic line-of-sight (LOS) velocities (${\sim}90\%$,
J+24).\footnotemark{} However, we find that our conclusions are unchanged when
limiting samples to only the brightest stars. Altogether, the J+24 method
provides a high-quality, low-contamination sample of dwarf galaxy candidate
member stars, which we will now investigate in further detail.

\footnotetext{This
  would indicate that the J+24 model probabilities are miss-calibrated.
  However, most LOS surveys of dwarf galaxies select brighter stars
  (which have better \emph{Gaia} measurements) and likely members,
  complicating the interpretation of this purity estimate.}

Fig.~\ref{fig:observed_density_profiles} compares the density
profiles of Scl, UMi, and eight other classical dwarf
galaxies.\footnote{
    We define classicals to be non-Magellanic dwarfs brighter than
    $Mv = -7.7$, i.e., Sagittarius, Fornax, Leo I, Sculptor, Antlia II, Leo II,
    Carina, Draco, Ursa Minor, Canes Venatici I, Sextans I, and Crater II. Of
    these dwarfs, we exclude Antlia II, due to the extremely high background,
    and Sagittarius, which was not included in J+24. 
}
All of the classical dwarfs appear to be well
described by an exponential profile in the inner regions.\footnote{
    Comparing
    density profiles of dwarf galaxies is complicated by variations in the
    effective depth between galaxies. However, the deviations between Scl,
    UMi, and other galaxies are apparent even where the data is complete
    across all dwarfs.} 
In the outer regions, however, Scl and Ursa
Minor deviate and show a clear outer excess over an exponential law
(solid black line). These galaxies are better fit by a Plummer law
(dashed black line). The deviation from an exponential grows outwards,
and at \(\sim 8 R_h\), may reach 2 orders of magnitude.

\begin{figure}
    \centering
    \includegraphics{figures/classical_dwarf_profiles.pdf}
    \caption{
        The density profiles of
        Sculptor (orange squares), Ursa Minor (red triangles), and other
        classical dwarfs (blue lines) compared to 2D exponential (solid black
        line) and Plummer density profiles (dashed black line). Density
        profiles are plotted in terms of the logarithm of elliptical radius
        scaled by the half-light radius from \citet{munoz+2018}, and the
        logarithmic stellar density scaled by the density at half-light
        radius. The residuals (lower panel) are with respect to the 2D
        exponential. Sculptor and Ursa Minor have an excess of stars in the
        outer regions (past \(\log R/R_h \sim 0.3\)) compared with other
        classical dwarfs.
}

    \label{fig:observed_density_profiles}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

\subsection{Milky Way potential and reference frame}\label{milky-way-potential}
To convert observed positions and velocities to Galactocentric
Cartesian coordinates, we use the Astropy v4 right-handed Galactocentric frame
\citep{astropycollaboration+2022}. We assume $x$ is the direction from the sun to
the Galactic centre, $y$ is the direction of the motion of the Local
Standard of Rest, and $z$ is the direction perpendicular to the
Galactic plane. 
% The coordinate frame is also right-handed, such that the
% \(z\)-angular momentum of the sun is negative (since the sun is at
% \(x<0\)). In this frame, the solar position is
% \([-8.122 \pm 0.033, 0, 0.0208 \pm 0.003]\, \kpc\)
% \citep{gravitycollaboration+2018, bennett+bovy2019} and the solar
% velocity is \(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\)
% km/s
% \citep{reid+brunthaler2004, drimmel+poggio2018, gravitycollaboration+2018}.
% The uncertainties in this reference frame are typically smaller than the
% uncertainties on a dwarf galaxy's distance and tangential velocity.

For our MW-only models, we adopt the potential from \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
The Potential consists of a \citet{hernquist1990} bulge, a \citet{miyamoto+nagai1975} thin and thick disk, and a NFW \citep{NFW1996, NFW1997} dark matter halo.
A NFW density profile is defined as,
\begin{equation}\label{eq:nfw}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation}
where $r_s$ and $\rho_s$ are the scale mass and scale density. We later parameterize the NFW halo instead in terms of $\rmax$ and $\vmax$, the maximum circular velocity and radius of maximum circular velocity.  
In terms of these parameters, \(\rmax = 43.7\,\)kpc and \(\vmax = 191\,\kms\). Equivalently, the dark matter halo has a virial mass\footnotemark{} \(M_{200} = 126.6\times 10^{10}\,\Mo\) and \(r_s=20.2\,\kpc\).
\footnotetext{The virial mass is the mass enclosed within the region containing a mean density 200 times the Universe's critical density}
%The Galactic bulge is described by a \citet{hernquist1990} potential,
%\begin{equation}{
%    \Phi(r) = - \frac{GM}{r + a},
%}\end{equation} 
%where \(a=1.3\,{\rm kpc}\) is the scale radius and
%\(M=2.1 \times 10^{10}\,\Mo\) is the total mass. The thin and thick
%disks are represented with the \citet{miyamoto+nagai1975} cylindrical
%potential, 
%\begin{equation}{
%    \Phi(R, z) = \frac{-GM}{\sqrt{R^2 + \left(a + \sqrt{z^2 + b^2}\right)^{2}}},
%}\end{equation} 
%where \(a\) is the disc radial scale length, \(b\) is
%the scale height, and \(M\) is the total mass of the disk. For the thin
%disk, \(a=3.944\,\)kpc, \(b=0.311\,\)kpc, and
%\(M=3.944\times10^{10}\,\)M\(_\odot\). For the thick disk,
%\(a=4.4\,\)kpc, \(b=0.92\,\)kpc, and \(M=2\times10^{10}\,\)M\(_\odot\).

For our MW and LMC model, we use the \texttt{L3M11} model from
\citet{vasiliev2024}. The \texttt{L3M11} potential is an evolving
multipole approximation of an N-body simulation including a live MW and
LMC dark matter halo. The potential includes a static MW bulge and disk,
evolving MW and LMC halos, and the MW reflex motion. In their
simulation, the MW was initially a NFW halo with \(r_s=16.5\,\)kpc and
\(M_{\rm 200}= 98.4\times10^{10}\Mo\), and the LMC a NFW halo with
\(r_s=11.7\) and \(M_{200} = 24.6 \times 10^{10} \Mo\). The total
\texttt{L3M11} MW mass is lighter than our initial \citet{EP2020}
potential.

@@

\begin{table}[t]
\centering
\caption{The initial conditions for models presented in this work. The observables record the intended final position of the N-body model derived from point-mass orbit quantiles. Instead, the initial position and velocity represent the initial conditions for the actual N-body model. The \smallperi{} orbits represents the $3\sigma$-smallest-pericentre orbit, which we use to provide an upper limit on tidal effects}
\label{tbl:orbit_ics}
\begin{tabular}{llll}
\hline
Property & Scl: MW& Scl: MW+LMC & UMi: MW\\
\hline
$\vmax / \kms$ & 31 & 25 & 38 \\
$\rmax / \kpc$ & 3.2 & 2.5 & 4.0 \\
$R_{h, \rm exp}$ & 0.168 & 0.185 & 0.168 \\
$R_{h, \rm Plummer}$ & 0.20 & 0.20  & 0.20 \\
\hline
distance $/$ kpc & 82.6 & 73.1 & 64.6\\
$\mu_{\alpha*} / \masyr$ & 0.134 & 0.137 & -0.158\\
$\mu_\delta / \masyr$ & -0.198 & -0.156 & 0.050\\
$v_{\rm los}$ $/$ $\kms$ & 111.2 & 111.2 & -245.75\\
\hline
$t_i / \Gyr$ & -9.17 & -2.0 & -9.67\\
${x}_{i} / \kpc$ & -2.49 & 4.30 & -17.40\\
${y}_{i} / \kpc$ & -42.78 & 138.89 & 74.51\\
${z}_{i} / \kpc$ & 86.10 & 125.88 & 21.34\\
$\V_{x,\,i} / \kms$ & -20.56 & 6.89 & 14.27\\
$\V_{y,\,i} / \kms$ & -114.83 & -56.83 & 48.62\\
$\V_{z,\,i} / \kms$ & -57.29 & 52.09 & -114.08\\
\hline
\end{tabular}
\end{table}

To explore the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables (assuming Normal distributions
and the values in Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}). We
integrate
each sampled position/velocity back in time for 10 Gyr using \agama{}
\citep{agama}. Dynamical friction is not expected to impact orbits
substantially, so we assume a single point-mass particle for the backwards
integration.
Fig.~\ref{fig:orbits} shows 100 samples of these orbits in both potentials (MW only and MW+LMC), and the orbits of our N-body models.
